---
title: "core syntax notes"
output: pdf_document
---

# Core syntax, basic data handling in tidyverse

## Materials needed

-   `library(tidyverse)`
-   `library(dbplyr)`

## Rows

First, load tidyverse:

```{r}
library(tidyverse)
```

Then, we need to load some data. We are going to load the flights data. This is what we are going to be using.

```{r}
library(nycflights13)
flights
```

### `filter()`

Next we can use the `filter()` command to filter out some of the data. Let’s say we want to filter out departure delay times greater than 1:

```{r}
flights %>%
  filter(dep_delay > 1)
```

Or we can also add more than one filter. Let’s say we want to filter the depature delays > 1 AND are coming from the JFK airport. 

```{r}
flights %>%
  filter(dep_delay > 1 & origin == "JFK")
```

You can also use “or” reasoning so a condition is **either one or the other.** Use the **straight line (\|)** for the “or” reasoning. Note that this yields more results than the and reasoning.

```{r}
flights %>%
  filter(dep_delay > 1 | origin == "JFK") # either has a departure delay > 1 OR is coming from JFK
```

### `arrange()`

the `arrange()` command changes the order of the rows based on the value of the columns. For example, this one below arranges the year, month, and day by departure delay. 

```{r}
flights %>%
  arrange(year, month, day, dep_delay)
```

In this code below, the results are arranged by descending reaction time.

```{r}
flights %>%
  arrange(desc(dep_delay))
```

### `distinct()` and `count()`

`distinct` finds all the unique rows in a data set and then primarily operates on those rows. 
```{r}
flights %>%
  distinct() # finds any duplicate values (if any)
# looks like there are none!
```

Next we are going to find all the unique pairs of origin and destination.

```{r}
flights %>%
  distinct(origin, dest)
```

Alternatively, **you can keep all the other columns using `.keep_all=TRUE`**.

```{r}
flights %>%
  distinct(origin, dest, .keep_all=TRUE)
```

You can also use the `count()` command to find the number of occurrences instead.

```{r}
flights %>%
  count(origin, dest, sort=TRUE)
```
## Columns
There are four functions we can use to change columns without affecting the rows. 
- `mutate()`
- `select()`
- `rename()`
- `relocate`

### `mutate()`
First I’m going to use mutate. Mutate is used when you want to add new columns calculated from other ones. 
I want to add three new columns: `gain`, `hours`, and `gain_per_hr`. 
```{r}
flights %>%
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used" # indicates that I want to keep the columns I used in the function. 
  )
```
We can use `.before` and `.after` to put the columns before or after a specific column. 
```{r}
flights %>%
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .before = 1
  )
```
```{r}
flights %>%
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .after = 1
  )
```
