---
title: "core syntax notes"
output: pdf_document
---

# Core syntax, basic data handling in tidyverse

## Materials needed

-   `library(tidyverse)`
-   `library(dbplyr)`

## Rows

First, load tidyverse:

```{r}
library(tidyverse)
```

Then, we need to load some data. We are going to load the sleep study data. This is what we are going to be using.

```{r}
library(lme4)
data(sleepstudy)
head(sleepstudy)
```

### `filter()` command

Next we can use the `filter()` command to filter out some of the data. Let’s say we want to filter out reaction times greater than 200ms:

```{r}
sleepstudy %>%
  filter(Reaction > 200)
```

Or we can also filter by days. Let’s pick day six for example. **Use the ampersand (&) to make more than one filter active.**

```{r}
sleepstudy %>%
  filter(Reaction > 200 & Days == 6)
```

You can also use “or” reasoning so a condition is **either one or the other.** Use the **straight line (\|)** for the “or” reasoning. Note that this yields more results than the and reasoning.

```{r}
sleepstudy %>%
  filter(Reaction > 200 | Days == 6) # either greater than 200ms OR is day 6
```

### `arrange()` command

the `arrange()` command changes the order of the rows based on the value of the columns. For example, this one below **arranges increasing reaction time by subject.**

```{r}
sleepstudy %>%
  arrange(Reaction, Subject)
```

In this code below, the results are arranged by descending reaction time.

```{r}
sleepstudy %>%
  arrange(desc(Reaction))
```

### `distinct()` and `count()` command

`distinct` finds all the unique rows in a data set and then primarily operates on those rows. I’m actually going to be using a different data set. This is the Dark Triad data set.

```{r}
SD3_data <- read_csv("C://Users//thomp//remote//r-reference//downloaded_datasets//SD3_data.csv")
print(SD3_data)
```

Next, we get rid of the duplicates (if any)

```{r}
SD3_data %>%
  distinct()
```

Next we are going to find all the unique pairs in country (V28) and P1 (“I like gettingto get revenge on authorities”).

```{r}
SD3_data %>%
  distinct(P1, country)
```

Alternatively, **you can keep all the other columns using `.keep_all=TRUE`**.

```{r}
SD3_data %>%
  distinct(P1, country, .keep_all=TRUE)
```

You can also use the `count()` command to find the number of occurrences instead.

```{r}
SD3_data %>%
  count(P1, country, sort=TRUE)
```
## Columns

