---
title: "R reference notes"
output: html_document
---

# Core syntax, basic data handling in tidyverse

## Materials needed

-   `library(tidyverse)`
-   `library(dbplyr)`

## Rows

First, load tidyverse:

```{r}
library(tidyverse)
```

Then, we need to load some data. We are going to load the flights data. This is what we are going to be using.

```{r}
library(nycflights13)
head(flights)
```

### `filter()`

Next we can use the `filter()` command to filter out some of the data. Let’s say we want to filter out departure delay times greater than 1:

```{r}
flights_1 <- flights %>%
  filter(dep_delay > 1)
```

Or we can also add more than one filter. Let’s say we want to filter the depature delays \> 1 AND are coming from the JFK airport.

```{r}
flights %>%
  filter(dep_delay > 1 & origin == "JFK")
```

You can also use “or” reasoning so a condition is **either one or the other.** Use the **straight line (\|)** for the “or” reasoning. Note that this yields more results than the and reasoning.

```{r}
flights %>%
  filter(dep_delay > 1 | origin == "JFK") # either has a departure delay > 1 OR is coming from JFK
```

### `arrange()`

the `arrange()` command changes the order of the rows based on the value of the columns. For example, this one below arranges the year, month, and day by departure delay.

```{r}
flights %>%
  arrange(year, month, day, dep_delay)
```

In this code below, the results are arranged by descending reaction time.

```{r}
flights %>%
  arrange(desc(dep_delay))
```

### `distinct()` and `count()`

`distinct` finds all the unique rows in a data set and then primarily operates on those rows.

```{r}
flights %>%
  distinct() # finds any duplicate values (if any)
# looks like there are none!
```

Next we are going to find all the unique pairs of origin and destination.

```{r}
flights %>%
  distinct(origin, dest)
```

Alternatively, **you can keep all the other columns using `.keep_all=TRUE`**.

```{r}
flights %>%
  distinct(origin, dest, .keep_all=TRUE)
```

You can also use the `count()` command to find the number of occurrences instead.

```{r}
flights %>%
  count(origin, dest, sort=TRUE)
```

## Columns

There are four functions we can use to change columns without affecting the rows.

-   `mutate()`

-   `select()`

-   `rename()`

-   `relocate`

### `mutate()`

First I’m going to use mutate. Mutate is used when you want to add new columns calculated from other ones. I want to add three new columns: `gain`, `hours`, and `gain_per_hr`.

```{r}
flights %>%
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used" # indicates that I want to keep the columns I used in the function. 
  )
```

We can use `.before` and `.after` to put the columns before or after a specific column.

```{r}
flights %>%
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .before = 1
  )
```

```{r}
flights %>%
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .after = 1
  )
```

### `select()`

The next one we are going to go over is select. Select allows us to… well, select specific columns.

```{r}
flights %>%
  select(year, month, day)
```

Selecting columns between specific columns using a colon.

```{r}
flights %>%
  select(year:day)
```

Select all columns EXCEPT for those from year to day:

```{r}
flights %>%
  select(!year:day)
```

Select columns that are just characters:

```{r}
flights %>%
  select(where(is.character))
```

Here are some other helper functions:

-   `starts_with("abc")`: matches names that begin with “abc”.

-   `ends_with("xyz")`: matches names that end with “xyz”.

-   `contains("ijk")`: matches names that contain “ijk”.

-   `num_range("x", 1:3)`: matches x1, x2 and x3.

You can also rename variables using the equal sign:

```{r}
flights %>%
  select(tail_num = tailnum)
```

### `rename()`

Rename is very straightforward. You can rename your variables within the table using the command with the equal sign.

```{r}
flights %>%
  rename(tail_num = tailnum)
```

If you have a bunch of inconsistently named columns and it would be painful to fix them all by hand, check out [`janitor::clean_names()`](https://sfirke.github.io/janitor/reference/clean_names.html) which provides some useful automated cleaning.

### `relocate()`

You can use relocate to, well, relocate variables within the table.

```{r}
flights %>%
  relocate(time_hour, air_time)
```

You can also specify where to relocate them.

```{r}
flights %>%
  relocate(
    starts_with("J"), # relocates anything starting with J
    .after = dep_time
  )
```
